A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-03-30 analysis/paper/paper.Rmd
Minor edits

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform a refined method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @creel2022; @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement data is based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5]. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)). The SLIPs indicate the isolation of the basin from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range. This is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the standard squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline. This can then be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:

@@

Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shore-line. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.

>>>>

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform an improved method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999; and recent compilation by @creel2022], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement data is based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5]. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)). The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range. This is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the standard squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phases. In the case where there are multiple dates believed to belong to a single phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline. This can then be combined with the elevation of the site and the shoreline displacement to provide a method for shoreline dating that takes this distance into account:

@@

Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.
